Bayesian regression facilitates quantitative modelling of cell metabolism

### 2023-10-31 paper/main.qmd
Update paper

>>>>

Marginal posterior distributions from our main model run. (**Top left**)
Comparison of posterior predictive intervals with simulated flux measurement
values. All the flux measurements are within the predictive intervals,
indicating a good fit. (**Top right**) Comparison of posterior predictive
intervals with simulated concentration measurement values. These also show a
good fit. (**Bottom left**) Pairwise marginal posterior distribution for two
uncorrelated parameters, namely $K_m^{AHC1,hcys-L}$ and $K_m^{AHC1_ahcys}$.
(**Bottom right**) Pairwise marginal posterior distribution for two correlated
parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.

### 2023-10-31 paper/main.tex
Update paper

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

Posterior predictive checking indicated that our model achieved a good
fit to the simulated reaction and metabolite concentration measurements,
as shown by the graphs in the top row of figure 3. Note that the fit was
good for both training and validation measurements.

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
figure 3; in this case the marginal distribution of the two parameters
is roughly banana shaped. More commonly, however, two-dimensional pair
plots were insufficient to reveal the underlying correlation structure,
as seen in the bottom-right plot in figure 3. This does not mean that
the parameters were uncorrelated, but rather that the correlations
involve more than two parameters.

@@

@@

To demonstrate how Maud's output can be used to yield useful metabolic
insights we used the results of our case study to explain why the flux
of the enzyme \(GNMT\) is higher in condition 1 than in condition 2.
GNMT is an irreversible enzyme that is homotropically activated by its
substrate, competitively inhibited by its product and heterotropically
inhibited. The diverse regulation makes it the ideal test case to
elucidate regulatory changes.

Figure~\ref{fig-decomposition} shows the regulatory description of GNMT,
according to the results our main case study analysis. Each curve shows
the marginal posterior distribution of the ratio of the corresponding
regulatory component in condition 1 compared with condition 2. A
positive value indicates that the component was increased in condition 1
relative to condition 2, with 0 indicating no difference. The
probability, according to our model, of the component acting in each
direction is given by the relative area under the curve on each side of
the 0 point.

Our model correctly inferred that saturation and allosteric effects were
the main drivers of regulation between the two conditions in this case,
with the curves for each component aligning with the ground truth shown
in red.

Importantly, this form of analysis takes into account all modelled
sources of uncertainty, including uncertainty about the true values of
the flux in each condition, and propagates this uncertainty into the
final conclusion. Our result shows that Maud could be used in this
realistic case not only to provide a user with an explanation for an
observed difference in fluxes, but also a reasonable judgement as to the
explanation's robustness.

@@

\caption{\label{fig-decomposition}Illustration of how analysing a system
with Maud can yield actionable insights about the underlying metabolic
network. (\textbf{Top Left}) A schematic of the regulatory interactions
associated with the enzyme \(GNMT1\). Dashed green lines represents
allosteric activation, dashed red lines indicate allosteric inhibition
and solid red lines represent competitive inhibition. (\textbf{Bottom
Left}) Comparison of marginal posterior distributions for \(GNMT\) flux
in conditions 1 and 2. (\textbf{Right}) Log-scale ratios of the
regulatory elements defined in equation \eqref{eq-decomposition}. Note
that the \(Reversibility\) and \(k_cat\) components are excluded: this
is because this reaction was modelled as irreversible, and \(k_{cat}\)
was modelled as constant across conditions. These plots identify why
flux in condition 2 is higher than in condition 1: the flux increase is
due to allostery and saturation with no control from enzyme
concentration changes.}

>>>>

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, enhanced, boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Posterior predictive checking indicated that our model achieved a good
fit to the simulated reaction and metabolite concentration measurements,
as shown by the graphs in the top row of figure
Figure~\ref{fig-posterior}.

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
figure Figure~\ref{fig-posterior}; in this case the marginal
distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to
reveal the underlying correlation structure, as seen in the bottom-right
plot in figure Figure~\ref{fig-posterior}. This does not mean that the
parameters were uncorrelated, but rather that the correlations involve
more than two parameters.

@@

\includegraphics{../figures/posterior.png}

\caption{\label{fig-posterior}Marginal posterior distributions from our
main model run. (\textbf{Top left}) Comparison of posterior predictive
intervals with simulated flux measurement values. All the flux
measurements are within the predictive intervals, indicating a good fit.
(\textbf{Top right}) Comparison of posterior predictive intervals with
simulated concentration measurement values. These also show a good fit.
(\textbf{Bottom left}) Pairwise marginal posterior distribution for two
uncorrelated parameters, namely \(K_m^{AHC1,hcys-L}\) and
\(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
posterior distribution for two correlated parameters, namely
\(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}

@@

To demonstrate how Maud's output can be used to yield useful metabolic
insights we used the results of our case study to explain why the flux
of the enzyme \(GNMT\) is higher in dataset 1 than in dataset 12. GNMT
is an irreversible enzyme that is homotropically activated by its
substrate, competitively inhibited by its product and heterotropically
inhibited by 5,10- methylenetetrahydrofolate (mlthf). The diverse
regulation makes it the ideal test case to elucidate regulatory changes.

Figure~\ref{fig-decomposition} shows the regulatory description of GNMT,
according to the results our main case study analysis. Each curve shows
the marginal posterior distribution of the ratio of the corresponding
regulatory component in dataset 1 compared with dataset 12. A positive
value indicates that the component was increased in dataset 1 relative
to dataset 12, with 0 indicating no difference. The probability,
according to our model, of the component acting in each direction is
given by the relative area under the curve on each side of the 0 point.

Our model correctly inferred that saturation and allosteric effects were
the main drivers of regulation between the two datasets in this case,
with the curves for each component aligning with the ground truth shown
in red.

Importantly, this form of analysis takes into account all modelled
sources of uncertainty, including uncertainty about the true values of
the flux in each dataset, and propagates this uncertainty into the final
conclusion. Our result shows that Maud could be used in this realistic
case not only to provide a user with an explanation for an observed
difference in fluxes, but also a reasonable judgement as to the
explanation's robustness.

@@

\caption{\label{fig-decomposition}Illustration of how analysing a system
with Maud can yield actionable insights about the underlying metabolic
network. (\textbf{Top Left}) A schematic of the regulatory interactions
associated with the enzyme \(GNMT1\). Dashed green lines represents
allosteric activation, dashed red lines indicate allosteric inhibition
and solid red lines represent competitive inhibition. (\textbf{Bottom
Left}) Comparison of marginal posterior distributions for \(GNMT\) flux
in datasets 1 and 2. (\textbf{Right}) Log-scale ratios of the regulatory
elements defined in equation \eqref{eq-decomposition}. Note that the
\(Reversibility\) and \(k_cat\) components are excluded: this is because
this reaction was modelled as irreversible, and \(k_{cat}\) was modelled
as constant across datasets. These plots identify why flux in dataset 12
is higher than in dataset 1: the flux increase is due to allostery and
saturation with no control from enzyme concentration changes.}
